Does differentiation beget differentiation?

### 2022-07-02 Preprint-Differentiation.Rmd
Uploaded the newest version of .Rmd
This includes the newest version of the .Rmd file necessary for replication.

---
title: "Does differentiation beget differentiation?"
subtitle: Examining the link between exposure to and support for differentiated integration
author: "Martin Moland^[ARENA, Centre for European Studies]"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: pdf_document
abstract: "Questions about the public legitimacy of the EU's current flexible integration are becoming increasingly important in the public opinion literature. However, we still know little about how exposure to past differentiation shapes public preferences for a more differentiated EU. Combining survey data from 2020-21 with data detailing EU countries’ differentiations from the EU treaties I show that citizens of countries previously exposed to differentiations from the EU’s legal order are more likely to want a less uniform EU integration in the future. This has implications both for the debate about the shape of the future EU and the broader literature investigating public support for transnational cooperation."
csl: european-journal-of-political-research.csl
bibliography: bib_file.bib
header-includes: 
  - \usepackage{setspace}\doublespacing
  - \usepackage{float}
fontsize: 12pt

@@

My findings are relevant to the discussion about the EU's future that Brexit gave rise to [@european_commission_white_2017]: As differentiated integration has become a more important tool of EU governance, especially given the EU’s increasing integration of for instance monetary and fiscal policy [@schimmelfennig_ever_2020; @genschel_differentiated_2013], greater exposure to it among member state citizens may make it more difficult for the EU’s institutions to consolidate support for the project of further integration. The article also makes an important contribution to the emerging literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009], by showing that implementation of differentiations from international law can potentially increase public demand for future divergences from the same legal order. Lastly the paper shows that differentiation, favoured as a tool for mitigating the contestation provoked by the EU’s increasing integration of particularly salient policy areas [@genschel_market_2018], may complicate the EU’s progress towards its goal of “ever closer Union”.

@@

This literature, in other words, mostly focuses on predictors found at the level of individuals. However, popular attitudes towards both the EU and international cooperation more broadly are often shaped by elite cues and context [@dellmuth_elite_2021; @edwards_public_2009; @harteveld_europe_2013; @telle_elite-mass_2022-1]. One under-explored source of such support is the large variation in differentiated integration found across the EU [@schimmelfennig_instrumental_2014]: As either externally imposed or voluntarily chosen differentiation may be either in or out of step with popular attitudes towards how the EU should look, past differentiation may face a popular response in the form of lesser or greater support for differentiation [@busemeyer_beyond_2021; @wlezien_public_1995]. While previous literature has investigated how this feedback mechanism plays out in the case of EU policies [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], this paper will instead investigate how differentiations from the EU's primary legal framework shapes demand for a differently configured polity.

The effect of exposure to differentiation may also be compounded by who originally initiated differentiations, and the circumstances under which it happened. An interesting test of this assumption is support for differentiation in Central and Eastern Europe. Here citizens were exempt from the right to move across EU borders to seek work for several years after their countries acceded to the EU. Differentiation was thus externally imposed as a precondition for EU membership [@cianciara_differentiated_2014; @schimmelfennig_eu_2014]. This was done because existing member states wanted to shield themselves from the economic destabilization that they feared would follow from the EU accession of countries like Poland, Estonia, and Hungary [@schimmelfennig_eu_2014]. This could easily be seen by those excluded from these policies as a cartelization of the benefits of EU integration [@lord_autonomy_2021 547]. Citizens of Central and Eastern Europe could thus respond to the involuntary nature of the exposure they have had to differentiated integration by expressing lower support for it in the future.

However, another causal mechanism could explain lower levels of support for differentiated integration in Central and Eastern Europe without relying on a path dependency argument: Central and Eastern European politicians have framed differentiated integration as a path towards an EU consisting of a powerful core and a less powerful periphery. This frame was prominent during the sovereign debt crisis [@cianciara_differentiated_2014 178-179]. For example, Polish politicians opposed further integration of the Eurozone, a clear example of differentiated integration, because they feared that the Eastern member states would become less influential in economic policy-making due to being non-Eurozone countries. Such cues have been known to shape popular attitudes towards integration, particularly where citizens have little knowledge of the relevant policies. It may thus help explain possible opposition to differentiated integration [@steenbergen_whos_2007; @telle_elite-mass_2022-1].

@@

Evidence of this is found in how the renewed Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy on the part of those identifying as Eurosceptic [@schraff_does_2020]. This suggests that citizens may respond positively to differentiation, and could indicate that they would be favourably inclined towards further differentiation. This trend might also be particularly strong among Eurosceptics in these countries: These citizens may, because of their countries’ relatively small sizes, be more likely to see differentiation as a viable second-best alternative to the European disintegration they would otherwise have favoured. This could make them more likely than supporters of EU membership to favour further differentiation. I formulate the following hypotheses:  
*H3: The positive effect of exposure to differentiated integration will be greater for citizens of Sweden and Denmark than in the entire EU.*  
*H3a: The increase in support attributable to increased exposure to differentiated integration will be greater for Scandinavian Eurosceptics than supporters of EU membership.*

@@

# Data and methods
The hypotheses will be investigated using multilevel linear models and data from Yougov surveys fielded in 2020-21 [@hemerijck_eui-yougov_2021]. I combine these data with the expanded EUDIFF1 dataset [@schimmelfennig_eudiff1_2022]. This dataset features all distinct differentiations from the EU treaties, showing for each country the number of articles in which differentiation was given, the years they were in force and the policy areas covered. The surveys, on the other hand, ask a range of questions related to European integration. Two of these ask about preferences for differentiated integration. The first question asks people whether they want an EU where countries may opt out of certain policy integration. This may be termed a constitutionally differentiated EU. The second asks whether the EU should allow countries to arrive at the same level of integration at different speeds when they are ready. This would be an instrumentally differentiated EU. This article focuses on support for constitutional differentiated integration, whose permanent segmentation represents arguably the strongest departure from the EU's current goal of continuously closer integration. Because it allows for both fragmentation with implications for each country's power within the union as well as greater space for national autonomy, attitudes towards  differentiation are likely to be shaped by both sovereigntist discourses and concerns about the relative strength of one's country within the EU.

>>>>

---
title: "Does differentiation beget differentiation?"
subtitle: Examining the link between exposure to and support for differentiated integration
author: "Martin Moland^[ARENA, Centre for European Studies]"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: 
  pdf_document:
    keep_md: true
abstract: "Questions about the public legitimacy of the EU's current flexible integration are becoming increasingly important in the public opinion literature. However, we still know little about how exposure to past differentiation shapes public preferences for a more differentiated EU. Combining survey data from 2020-21 with data detailing EU countries’ differentiations from the EU treaties I show that citizens of countries previously exposed to differentiations from the EU’s legal order are more likely to want a less uniform EU integration in the future. This has implications both for the debate about the shape of the future EU and the broader literature investigating public support for transnational cooperation."
csl: european-journal-of-political-research.csl
bibliography: bib_file.bib
header-includes: 
  - \usepackage{setspace}\doublespacing
  - \usepackage{float}
fontsize: 12pt

@@

My findings are relevant to the discussion about the EU's future spawned by Brexit [@european_commission_white_2017]: As differentiated integration has become a more important tool of EU governance, especially given the EU’s increasing integration of for instance monetary and fiscal policy [@schimmelfennig_ever_2020; @genschel_differentiated_2013], greater exposure to it among member state citizens may make it more difficult for the EU’s institutions to consolidate support for the project of further integration. The article also makes an important contribution to the emerging literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009], by showing that implementation of differentiations from international law can potentially increase public demand for future divergences from the same legal order. Lastly the paper shows that differentiation, favoured as a tool for mitigating the contestation provoked by the EU’s increasing integration of particularly salient policy areas [@genschel_market_2018], may complicate the EU’s progress towards its goal of “ever closer Union”.

@@

This literature, in other words, mostly focuses on predictors found at the level of individuals. However, popular attitudes towards both the EU and international cooperation in general are often shaped by elite cueing and national contexts [@dellmuth_elite_2021; @edwards_public_2009; @harteveld_europe_2013; @telle_elite-mass_2022-1]. One under-explored source of such support is the large variation in differentiated integration found across the EU [@schimmelfennig_instrumental_2014]. Because differentiation may be either externally imposed or voluntarily chosen and more or less comprehensive than what citizens prefer, it may be out of step with popular preferences regarding the shape of the future EU. This may lead to a thermostatic response, in the form of either greater or lesser support for future differentiated integration [@busemeyer_beyond_2021; @wlezien_public_1995]. While previous literature has investigated how this feedback mechanism plays out at the level of policies [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], this paper will instead investigate how differentiations from the EU's primary legal framework shapes demand for a differently configured polity.

The effect of exposure to differentiation may also be compounded by who originally initiated it. An interesting test of this is to compare support for differentiation in Central and Eastern Europe to that found in the rest of the EU. Here citizens were exempt from the right to move across EU borders to seek work for several years after their countries acceded to the EU. Differentiation was thus externally imposed as a precondition for EU membership [@cianciara_differentiated_2014; @schimmelfennig_eu_2014]. This was done because existing member states wanted to shield themselves from the economic destabilization that they feared would follow from the EU accession of countries like Poland, Estonia, and Hungary [@schimmelfennig_eu_2014]. This could easily be seen by those excluded from these policies as a cartelization of the benefits of EU integration [@lord_autonomy_2021 547]. Citizens of Central and Eastern Europe could thus respond to the involuntary nature of the exposure they have had to differentiated integration by expressing lower support for it in the future.

However, another causal mechanism could explain lower levels of support for differentiated integration in Central and Eastern Europe: Central and Eastern European politicians have framed differentiated integration as a path towards an EU consisting of a powerful core and a less powerful periphery. This frame was prominent during the sovereign debt crisis [@cianciara_differentiated_2014 178-179]. For example, Polish politicians opposed further integration of the Eurozone, a prominent example of differentiated integration, because they feared that the Central and Eastern member states would become less influential in economic policy-making due to being non-Eurozone countries. Such cues have been known to shape popular attitudes towards integration, particularly where citizens have little knowledge of the relevant policies. It may thus help explain possible opposition to differentiated integration [@steenbergen_whos_2007; @telle_elite-mass_2022-1].

@@

This is evident in how renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy on the part of those identifying as Eurosceptic [@schraff_does_2020]. This suggests that citizens may respond positively to differentiation, and could indicate that they would be favourably inclined towards further differentiation. This trend might also be particularly strong among Eurosceptics in these countries: These citizens may, because of their countries’ relatively small sizes, be more likely to see differentiation as a viable second-best alternative to the European disintegration they would otherwise have favoured. This could make them more likely than supporters of EU membership to favour further differentiation. I formulate the following hypotheses:  
*H3: The positive effect of exposure to differentiated integration will be greater for citizens of Sweden and Denmark than in the entire EU.*  
*H3a: The increase in support attributable to increased exposure to differentiated integration will be greater for Scandinavian Eurosceptics than supporters of EU membership.*

@@

# Data and methods
The hypotheses will be investigated using multilevel linear models and data from Yougov surveys fielded in 2020-21 [@hemerijck_eui-yougov_2021]. I combine these data with the expanded EUDIFF1 dataset [@schimmelfennig_eudiff1_2022]. This dataset features all distinct differentiations from the EU treaties, showing for each country the number of articles to which differentiation was applied, the years they were in force and the policy areas covered. The surveys, on the other hand, ask a range of questions related to European integration. Two of these ask about preferences for differentiated integration. The first question asks people whether they want an EU where countries may opt out of certain policy integration. This may be termed a constitutionally differentiated EU. The second asks whether the EU should allow countries to arrive at the same level of integration at different speeds when they are ready. This would be an instrumentally differentiated EU. This article focuses on support for constitutional differentiated integration, whose permanent segmentation represents arguably the strongest departure from the EU's current goal of continuously closer integration. Because it allows for both fragmentation with implications for each country's power within the union as well as greater space for national autonomy, attitudes towards  differentiation are likely to be shaped by both sovereigntist discourses and concerns about the relative strength of one's country within the EU.
